---
title: "01_setup"
author: "Sglatt"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set-up for application of the Book of OHDSI (Observational Health Data Sciences and Informatics) (link: https://ohdsi.github.io/TheBookOfOhdsi/)

HADES (formally known as the OHDSI Methods Library) is a set of open source R packages for large scale analytics, including population characterization, population-level causal effect estimation, and patient-level prediction

The packages offer R functions that together can be used to perform an observation study from data to estimates and supporting statistics, figures, and tables. The packages interact directly with observational data in the Common Data Model (CDM), and are designed to support both large datasets and large numbers of analyses (e.g. for testing many hypotheses including control hypotheses, and testing many analyses design variations). For this purpose, each Method package includes functions for specifying and subsequently executing multiple analyses efficiently. HADES supports best practices for use of observational data as learned from previous and ongoing research, such as transparency, reproducibility, as well as measuring of the operating characteristics of methods in a particular context and subsequent empirical calibration of estimates produced by the methods. For more information about HADES’ design considerations, please refer to the HADES paper.

HADES paper: https://ebooks.iospress.nl/doi/10.3233/SHTI231108; 

HADES paper citation: Schuemie M, Reps J, Black A, Defalco F, Evans L, Fridgeirsson E, Gilbert JP, Knoll C, Lavallee M, Rao GA, Rijnbeek P, Sadowski K, Sena A, Swerdel J, Williams RD, Suchard M. Health-Analytics Data to Evidence Suite (HADES): Open-Source Software for Observational Research. Stud Health Technol Inform. 2024 Jan 25;310:966-970. doi: 10.3233/SHTI231108.


# Configure JAVA
```{r java}
Sys.setenv(JAVA_HOME = "C:/Program Files/Java/jdk-27")

Sys.getenv("JAVA_HOME")
system2("java", "-version")


file.exists(
  file.path(
    Sys.getenv("JAVA_HOME"),
    "bin",
    "server",
    "jvm.dll"
  )
)

if (!require("rJava")) {
  install.packages("rJava")
  require("rJava")
}

packageDescription("rJava")[c("Version", "Built")]
.jinit()
```

# HADES packages

Installing all at once:
```{r hades packages, eval=FALSE}
options(install.packages.compile.from.source = "never")

if (!require("remotes")) {
  install.packages("remotes")
  require("remotes")
}

install_github("ohdsi/Hades", upgrade = "always")
```