---
title: "02_exercises"
author: "Sglatt"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercises Book of OHDSI (Observational Health Data Sciences and Informatics) (link: https://ohdsi.github.io/TheBookOfOhdsi/)

Most HADES packages need to interact with a database. HADES `databaseConnector` makes these connections

`SqlRender` +  `DatabaseConnector` provide a way to query data in the CDM

The `Eunomia` package provides a simulated dataset in the CDM that will run inside the local R session

Importing a synthetic dataset from Eunomia via these packages and doing some exercises

Note: the exercises have a different API then the latest version I installed, so there's troubleshooting sprinkled in here

# Data and packages
```{r Packages }
# Data and packages
options(repos = c(
  CRAN = "https://cran.rstudio.com"
))

if (!require("Eunomia")) {
  install.packages("Eunomia")
  library(Eunomia)
}

install.packages(
  c("dbplyr", "DatabaseConnector"),
  type = "binary"
)
library(DatabaseConnector)

if (!require("SqlRender")) {
  install.packages("SqlRender")
  library(SqlRender)
}

# Get the connection details
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
connectionDetails

# Connect using the SQLite driver
connection <- connect(connectionDetails)

# Create the cohorts used in Chapter 10
Eunomia::createCohorts(connectionDetails)

# Chapter 10
library(CohortMethod)

# Chapter 11
install.packages(
  "FeatureExtraction",
  repos = c(
    "https://ohdsi.github.io/drat",
    "https://cran.rstudio.com"
  )
)
library(FeatureExtraction)

# Chapter 13
install.packages(
  "PatientLevelPrediction",
  repos = c(
    "https://ohdsi.github.io/drat",
    "https://cran.rstudio.com"
  )
)
library(PatientLevelPrediction)
```

# Folder for any output
```{r folder}
if (!dir.exists("02_output")) {
  dir.create("02_output")
}
```

# Exercises with the synthetic data 

## Chapter 9, SQL and R 
```{r ch 9 exercises}
# These exercises use SQL to answer simple questions about the OMOP CDM

# 1. How many people are in the database
sql <- "SELECT COUNT(*) AS person_count
FROM @cdm.person;"

renderTranslateQuerySql(connection, sql, cdm = "main") # 2694

# 2. How many people have at least one prescription of celecoxib?
sql <- "SELECT COUNT(DISTINCT(person_id)) AS person_count
FROM @cdm.drug_exposure
INNER JOIN @cdm.concept_ancestor
  ON drug_concept_id = descendant_concept_id
INNER JOIN @cdm.concept ingredient
  ON ancestor_concept_id = ingredient.concept_id
WHERE LOWER(ingredient.concept_name) = 'celecoxib'
  AND ingredient.concept_class_id = 'Ingredient'
  AND ingredient.standard_concept = 'S';"

renderTranslateQuerySql(connection, sql, cdm = "main") # 1844

# 3. How many diagnoses of gastrointestinal hemorrhage occur during exposure to celecoxib?
sql <- "SELECT COUNT(*) AS diagnose_count
FROM @cdm.drug_era
INNER JOIN @cdm.concept ingredient
  ON drug_concept_id = ingredient.concept_id
INNER JOIN @cdm.condition_occurrence
  ON condition_start_date >= drug_era_start_date
    AND condition_start_date <= drug_era_end_date
INNER JOIN @cdm.concept_ancestor
  ON condition_concept_id =descendant_concept_id
WHERE LOWER(ingredient.concept_name) = 'celecoxib'
  AND ingredient.concept_class_id = 'Ingredient'
  AND ingredient.standard_concept = 'S'
  AND ancestor_concept_id = 192671;"

renderTranslateQuerySql(connection, sql, cdm = "main") # 41
```

## Chapter 10, Defining a Cohort
```{r ch 10 exercises}
# Create a cohort for acute myocardial infarction (AMI) in the existing COHORT table, following these criteria:

# 1) An occurrence of a myocardial infarction diagnose (concept 4329847 “Myocardial infarction” and all of its descendants, excluding concept 314666 “Old myocardial infarction” and any of its descendants).

# 2) During an inpatient or ER visit (concepts 9201, 9203, and 262 for “Inpatient visit”, “Emergency Room Visit”, and “Emergency Room and Inpatient Visit”, respectively).


# For 1, find all AMI condition occurrences and store these in a temp table called “#diagnoses”
sql <- "SELECT person_id AS subject_id,
  condition_start_date AS cohort_start_date
INTO #diagnoses
FROM @cdm.condition_occurrence
WHERE condition_concept_id IN (
    SELECT descendant_concept_id
    FROM @cdm.concept_ancestor
    WHERE ancestor_concept_id = 4329847 -- Myocardial infarction
)
  AND condition_concept_id NOT IN (
    SELECT descendant_concept_id
    FROM @cdm.concept_ancestor
    WHERE ancestor_concept_id = 314666 -- Old myocardial infarction
);"

renderTranslateExecuteSql(connection, sql, cdm = "main")

# Then for 2, select  those that occur during an inpatient or ER visit, using some unique COHORT_DEFINITION_ID (here selected ‘1’)
sql <- "INSERT INTO @cdm.cohort (
  subject_id,
  cohort_start_date,
  cohort_definition_id
  )
SELECT subject_id,
  cohort_start_date,
  CAST (1 AS INT) AS cohort_definition_id
FROM #diagnoses
INNER JOIN @cdm.visit_occurrence
  ON subject_id = person_id
    AND cohort_start_date >= visit_start_date
    AND cohort_start_date <= visit_end_date
WHERE visit_concept_id IN (9201, 9203, 262); -- Inpatient or ER;"

renderTranslateExecuteSql(connection, sql, cdm = "main")
```

## Chapter 12, Population-Level Estimation
```{r ch. 12 exercises}
# Estimate the risk of gastrointestinal (GI) bleeding in new users of
# celecoxib compared with new users of diclofenac.


# celecoxib new-user cohort COHORT_DEFINITION_ID = 1
# diclofenac new-user cohort COHORT_DEFINITION_ID = 2
# GI bleed cohort COHORT_DEFINITION_ID = 3
# concept ID for celecoxib is 1118084
# diclofenac  ID for celecoxib 1124300,

# 1. Define covariates and extract CohortMethodData

nsaids <- c(1118084, 1124300) # celecoxib, diclofenac

covSettings <- createDefaultCovariateSettings(
  excludedCovariateConceptIds = nsaids,
  addDescendantsToExclude = TRUE)

# 2. Extract CohortMethodData from the CDM

# Troubleshooting...
# Different versions of CohortMethod use different function interfaces.
args(getDbCohortMethodData)
args(createGetDbCohortMethodDataArgs)

# Fix it
# In this version, covariateSettings has to be via createGetDbCohortMethodDataArgs()
getDbCohortMethodDataArgs <- createGetDbCohortMethodDataArgs(
  covariateSettings = covSettings
)

# Back to extracting the data
cmData <- getDbCohortMethodData(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  targetId = 1,
  comparatorId = 2,
  outcomeIds = 3,
  exposureDatabaseSchema = "main",
  exposureTable = "cohort",
  outcomeDatabaseSchema = "main",
  outcomeTable = "cohort",
  getDbCohortMethodDataArgs = getDbCohortMethodDataArgs
)

# Look at the data
summary(cmData)

# 3. Define the study population
# exclude subjects with a prior GI bleed
# start the risk window at cohort start
# follow subjects through the end of their cohort observation period

# Troubleshooting
args(createStudyPopulation)
args(createCreateStudyPopulationArgs)

# Study population
studyPop <- CohortMethod::createStudyPopulation(
  cohortMethodData = cmData,
  outcomeId = 3,
  createStudyPopulationArgs = CohortMethod::createCreateStudyPopulationArgs(
    removeSubjectsWithPriorOutcome = TRUE,
    riskWindowStart = 0,
    startAnchor = "cohort start",
    riskWindowEnd = 99999,
    endAnchor = "cohort end"
  )
)

# 4. Examine study-population attrition

# Draw the attrition/flowchart
drawAttritionDiagram(studyPop)

# 5. Fit an unadjusted Cox proportional hazards model

# Troubleshoot
args(fitOutcomeModel)
args(createFitOutcomeModelArgs)

cox_model <- fitOutcomeModel(
  population = studyPop,
  fitOutcomeModelArgs = createFitOutcomeModelArgs(
    modelType = "cox"
  )
)

# Look at the results
cox_model

# 6. Estimate the propensity score with covariates from above

ps <- createPs(cohortMethodData = cmData,
               population = studyPop)

# Plot the preference score distribution
plotPs(ps, showCountsLabel = TRUE, showAucLabel = TRUE)

# 7. Stratify on the propensity score and assess covariate balance

strataPop <- stratifyByPs(ps) # Divide subjects into propensity-score strata

# Now look at covariate balance before and after propensity-score stratification

bal <- computeCovariateBalance(strataPop, cmData)

# Plot the standardized covariate balance
plotCovariateBalanceScatterPlot(bal,
                                showCovariateCountLabel = TRUE,
                                showMaxLabel = TRUE,
                                beforeLabel = "Before stratification",
                                afterLabel = "After stratification")

# 8. Fit a Cox proportional hazards model using the propensity score strata
adjModel <- fitOutcomeModel(
  population = strataPop,
  fitOutcomeModelArgs = createFitOutcomeModelArgs(
    modelType = "cox",
    stratified = TRUE
  )
)

# Look at the results
adjModel
```

## Chapter 13, Patient-Level Prediction
```{r ch. 13 exercises}

# Predict which patients who are new users of NSAIDs will develop a
# gastrointestinal (GI) bleed during the following year.

# Study cohorts:
#   Target/exposure cohort: new NSAID users (COHORT_DEFINITION_ID = 4)
#   Outcome cohort: GI bleed (COHORT_DEFINITION_ID = 3)

# 1. Define the covariates and extract the PLP data

# Define the baseline characteristics that will be used as predictors

covSettings <- createCovariateSettings(
  useDemographicsGender = TRUE,
  useDemographicsAge = TRUE,
  useConditionGroupEraLongTerm = TRUE,
  useConditionGroupEraAnyTimePrior = TRUE,
  useDrugGroupEraLongTerm = TRUE,
  useDrugGroupEraAnyTimePrior = TRUE,
  useVisitConceptCountLongTerm = TRUE,
  longTermStartDays = -365,
  endDays = -1
)

# Troubleshooting; inspect the API for the installed PLP version

args(getPlpData)
args(createRestrictPlpDataSettings)
args(createDatabaseDetails)


# Create the database configuration used by getPlpData().

# cohortId = 4 identifies the NSAID new-user target cohort.
# outcomeIds = 3 identifies the GI bleed outcome cohort.

# The CDM, target cohort, and outcome cohort are all stored in the "main.cohort" table in this study

databaseDetails <- createDatabaseDetails(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  cdmDatabaseName = "main",
  cdmDatabaseId = 1,       # Replace with the actual CDM database ID if needed
  cohortDatabaseSchema = "main",
  cohortTable = "cohort",
  outcomeDatabaseSchema = "main",
  outcomeTable = "cohort",
  cohortId = 4,
  outcomeIds = 3
)


# Extract the PLP data from the CDM using the database configuration
# and the covariate definitions above.

plpData <- getPlpData(
  databaseDetails = databaseDetails,
  covariateSettings = covSettings
)


# Look at summary of the data

summary(plpData)


# 2. Define the prediction-study population
# The study population is defined around the NSAID cohort entry.

# Troubleshooting; look at the API used by the installed PLP version.

args(createStudyPopulation)
args(createStudyPopulationSettings)
formals(createStudyPopulation)


# createStudyPopulation() returns the actual study population
# createStudyPopulationSettings() defines the rules used to construct it.

plpStudyPop <- createStudyPopulation(
  plpData,
  3,
  createStudyPopulationSettings(
    washoutPeriod = 364,
    firstExposureOnly = FALSE,
    removeSubjectsWithPriorOutcome = TRUE,
    priorOutcomeLookback = 9999,
    riskWindowStart = 1,
    startAnchor = "cohort start",
    riskWindowEnd = 365,
    endAnchor = "cohort start",
    minTimeAtRisk = 364,
    requireTimeAtRisk = TRUE,
    includeAllOutcomes = TRUE
  )
)


# Check how many patients remain in the final prediction population.

nrow(plpStudyPop)


# 3. Specify the prediction model
# Use LASSO logistic regression.

lassoModel <- setLassoLogisticRegression(seed = 0)

# 4. Define the train/test split

# Troubleshooting; inspect the arguments accepted by runPlp() in the
# installed version of PatientLevelPrediction.

args(runPlp)

# Use a subject-level split 75/25 split
splitSettings <- createDefaultSplitSetting(
  type = "subject",
  testFraction = 0.25,
  trainFraction = 0.75,
  splitSeed = 123456,
  nfold = 2
)


# 5. Specify the population settings for runPlp()

# runPlp() expects the population settings
# So use the same study-population definition used above 

populationSettings <- createStudyPopulationSettings(
  washoutPeriod = 364,
  firstExposureOnly = FALSE,
  removeSubjectsWithPriorOutcome = TRUE,
  priorOutcomeLookback = 9999,
  riskWindowStart = 1,
  startAnchor = "cohort start",
  riskWindowEnd = 365,
  endAnchor = "cohort start",
  minTimeAtRisk = 364,
  requireTimeAtRisk = TRUE,
  includeAllOutcomes = TRUE
)


# 6. Run the LASSO prediction model

lassoResults <- runPlp(
  plpData = plpData,
  populationSettings = populationSettings,
  modelSettings = lassoModel,
  splitSettings = splitSettings,
  saveDirectory = "02_output"
)
```

